---
title: "Plasticity - Littorina offspring"
author: "Samuel"
date: "3/28/2020"
output:
  word_document: default
  html_document:
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

.packages = c("optparse", "dplyr", "tidyr", "rstan", "shinystan", "knitr", "kableExtra", "ggrepel",
              "flextable", "officer")
# Install CRAN packages (if not already installed)
.inst <- .packages %in% installed.packages()
if(length(.packages[!.inst]) > 0) install.packages(.packages[!.inst])
# Load packages into session
lapply(.packages, require, character.only=TRUE)
```

For now, the data and results belong to the island CZA and phenotype thickness.

```{r, echo=FALSE, warning=FALSE}
island = "CZA"
cz_phen = "mean_thickness"
dat_dir = paste0(island, "_off_SW/", island, "_off_final_data/")
dat_off = read.csv(file = paste0("../../",dat_dir, island, "_off_all_phenos_main_20200110.csv"))
dat_off = separate(data = dat_off, col = "snail_ID", into = c("pop", "ID"), sep = "_")
# table(nchar(as.character(dat_off$ID)))
dat_off[, "generation"] = 1
dat_off[which(nchar(as.character(dat_off$ID)) == 2), "generation"] = 0

dat_gen0 = dat_off[dat_off$generation==0, c("pop", "ID", cz_phen)]

dat_gen1 = dat_off[dat_off$generation==1, c("pop", "ID", cz_phen)]
diff_ypop = apply(X = dat_gen1[, c("pop", "ID")], MARGIN = 2,
                  FUN = function(x) !grepl(pattern = "x", x = x))
diff_ypop_idx = which(apply(diff_ypop, MARGIN = 1, FUN = sum)==2)
dat_gen1 = dat_gen1[diff_ypop_idx, ]

diff_pop = setdiff(dat_gen0$pop, dat_gen1$pop)
dat_gen0 = dat_gen0[dat_gen0$pop!=diff_pop,]

dat_gen0[, paste0("scaled_", cz_phen)] = (dat_gen0$mean_thickness - mean(dat_gen0$mean_thickness, na.rm = TRUE)) / sd(dat_gen0$mean_thickness, na.rm = TRUE)
x_meas = aggregate(x = dat_gen0[, paste0("scaled_", cz_phen)], by = list(pop = dat_gen0$pop),
                   FUN = function(x) c(mean = mean(x, na.rm = TRUE), sd = sd(x, na.rm = TRUE)))


dat_gen1[, paste0("scaled_", cz_phen)] = (dat_gen1$mean_thickness - mean(dat_gen0$mean_thickness, na.rm = TRUE)) / sd(dat_gen0$mean_thickness, na.rm = TRUE)
y_meas = aggregate(x = dat_gen1[, paste0("scaled_", cz_phen)], by = list(pop = dat_gen1$pop),
                   FUN = function(y) c(mean = mean(y, na.rm = TRUE), sd = sd(y, na.rm = TRUE)))
```

For each sample $i$ of population $j$ in both generations (parents and offspring), scaling was accomplished by substracting from the individual phenotype value the total mean phenotype and dividing by the total $SD$ phenotype of generation zero (parents). Total meaning generation zero (parents) across all populations.

$\hat{x_{ij}} = \frac{x_{ij} - \overline{x_{T}}}{\sigma_{x_{T}}}.$

$\hat{y_{ij}} = \frac{y_{ij} - \overline{x_{T}}}{\sigma_{x_{T}}}.$

where $x_{ij}$ and $y_{ij}$ is the phenotype value of sample $i$ of population $j$ in the parental and offspring generations, respectively. Total mean of parents across all populations is $\overline{x_T}$ and total $SD$ is referred as $\sigma_{x_T}$. Then, for each parental and offspring population, we calculated mean and $SD$ and used these measures as input for the regression model (Table 1; Fig. 1).


```{r tbl1, echo=FALSE}
tbl1 = data.frame(cbind(x_meas$x, y_meas$x))
if (length(strsplit(cz_phen, split = "_")[[1]])>1) {
  phen_h = strsplit(cz_phen, split = "_")[[1]][2]
} else {
  phen_h = cz_phen
}
var_h = c("mean", "sd")
gen_h = c("(P)", "(O)")

colnames(tbl1) = c(paste("scaled", var_h[1], phen_h, gen_h[1]), paste("scaled", var_h[2], phen_h, gen_h[1]),
                   paste("scaled", var_h[1], phen_h, gen_h[2]), paste("scaled", var_h[2], phen_h, gen_h[2]))
autofit(flextable(tbl1))
# autofit(flextable(as.data.frame(y_meas$x)))
```


```{r fig1, echo=FALSE, dpi=300, fig.height=2, fig.width=3, fig.cap="Figure 1. Scatter plot of the scaled data."}
ggplot(data = tbl1) +
    geom_abline(slope = 1, linetype="dashed") +
    geom_point(aes(x = tbl1[, 1], y = tbl1[, 3]),
               size=1) +
    geom_label_repel(aes(x = tbl1[, 1], y = tbl1[, 3], label = LETTERS[2:18]),
                     size = 2, box.padding = 0.2, label.padding = 0.15, segment.colour = "black",
                     min.segment.length = 0.1) +
    labs(x = "wild sample", y = "lab-reared sample") +
    theme(legend.position = "none",
          # plot.title = element_text(size = 19, hjust = 0.5),
          axis.title = element_text(size = 8),
          axis.text = element_text(size=5),
          axis.ticks = element_line(size = 0.5),
          panel.background = element_blank(),
          panel.border = element_rect(colour = "black", fill=NA, size=0.5),
          axis.line = element_line(size = 0.2, linetype = "solid",
                                   colour = "black"),
          panel.grid = element_line(colour = "gray70", size = 0.2))
```

\

We defined a regression model with measurement error in x an y using Stan.

```{stan output.var="err_in_var", eval=FALSE}
// The input data are vectors 'x' and 'sd_x' of length 'N'.
data {
  int<lower=0> N;
  vector[N] x;
  vector<lower=0>[N] sd_x;
}

// The parameters accepted by the model.
parameters {
  vector[N] x_lat;
  vector[N] y_lat;
  real alpha;
  real beta;
  real<lower=0> sigma;
}

transformed parameters {
  vector[N] mu_yhat = alpha + beta * x_lat;
}

// The model to be estimated. We model the output
// 'y_lat' to be normally distributed with mean 'mu_hat = alpha + beta * x_lat'
// and standard deviation 'sigma'.
model {
  //x_lat ~ normal(0, 10);
  alpha ~ normal(0, 5);
  beta ~ normal(1, 5);
  sigma ~ cauchy(0., 5);
  
  x ~ normal(x_lat, sd_x);
  y_lat ~ normal(mu_yhat, sigma);
}
```

There were warning messages when using $\sigma_{x_{T}}$ after scaling:

1: There were 59 divergent transitions after warmup. Increasing adapt_delta above 0.99 may help. See http://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup

2: There were 8059 transitions after warmup that exceeded the maximum treedepth. Increase max_treedepth above 10. See
http://mc-stan.org/misc/warnings.html#maximum-treedepth-exceeded

3: There were 4 chains where the estimated Bayesian Fraction of Missing Information was low. See
http://mc-stan.org/misc/warnings.html#bfmi-low

We can examine the Stan output to diagnose the above sampling problems (Fig. 2) and also look at the parameter estimate for _sigma_ (Table 2).

```{r fig2, echo=FALSE, fig.cap="Figure 2. Divergent transitions diagnosis."}
res_dir = paste0(island, "_off_SW/", island, "_off_results/")
pars = c("alpha", "beta", "sigma")
err_in_var = readRDS(paste0("../../", res_dir, "models/", island, "_err_in_var.rds"))
pairs(err_in_var, pars = pars, include = TRUE)
# pairs(err_in_var, pars = "x_lat", include = TRUE)
# pairs(err_in_var, pars = "y_lat", include = TRUE)
# pairs(err_in_var, pars = "mu_yhat", include = TRUE)
```

\

```{r tbl2, echo=FALSE}
stbl = read.csv(file = paste0("../../", res_dir, "tables/", island, "_err_in_var_stanfit.csv"))
pars_idx = which(as.character(stbl$X) %in% pars)
tbl2 = round(stbl[pars_idx, -1], 2)
autofit(flextable(data = cbind(par = stbl[pars_idx, 1], tbl2)))
```

\

Another alternative would be to try a mixed model with by population random effect.

\

Since we standardized the data we know that $\alpha$ should be centered near 0 and $\beta$ near 1 (Table 2). We would also expect the fitted values to fall onto the 1:1 line (Fig. 3).

```{r fig3, echo=FALSE, fig.cap="Figure 3. Scatter plot of the scaled data with fitted values."}
new_x_rng = range(stbl[grepl(pattern = "x_lat", x = as.character(stbl$X)), "mean"])
new_x = runif(n = 1000, min = new_x_rng[1], max = new_x_rng[2])
new_mu_yhat = stbl[pars_idx, "mean"][1] + stbl[pars_idx, "mean"][2] * new_x

fit_mu_yhat = stbl[grepl(pattern = "mu_yhat", x = as.character(stbl$X)), "mean"]

# sfitted = cbind(stbl[grepl(pattern = new_x, x = as.character(stbl$X)), c("X", "mean", "sd")],
#                 stbl[grepl(pattern = new_y, x = as.character(stbl$X)), c("X", "mean", "sd")])
# colnames(sfitted) = c("x_lat", "x_lat_mean", "x_lat_sd", "mu_yhat", "mu_yhat_mean", "mu_yhat_sd")

# sfitted = data.frame(x_lat_mean = new_x, mu_yhat_mean = new_mu_yhat)
sfitted = data.frame(fit_x_lat = stbl[grepl(pattern = "x_lat", x = as.character(stbl$X)), "mean"],
                     fit_mu_yhat = fit_mu_yhat)

ggplot(data = tbl1) +
    geom_abline(slope = 1, linetype="dashed") +
    geom_point(aes(x = tbl1[, 1], y = tbl1[, 3]),
               size=1) +
    geom_point(data = sfitted, aes(x = fit_x_lat, y = fit_mu_yhat), size=1, col="red") +
    # geom_line(data = sfitted, aes(x = x_lat_mean, y = mu_yhat_mean), size=0.5, col="red") +
    geom_label_repel(aes(x = tbl1[, 1], y = tbl1[, 3], label = LETTERS[2:18]),
                     size = 2, box.padding = 0.2, label.padding = 0.15, segment.colour = "black",
                     min.segment.length = 0.1) +
    labs(x = "wild sample", y = "lab-reared sample") +
    theme(legend.position = "none",
          # plot.title = element_text(size = 19, hjust = 0.5),
          axis.title = element_text(size = 8),
          axis.text = element_text(size=5),
          axis.ticks = element_line(size = 0.5),
          panel.background = element_blank(),
          panel.border = element_rect(colour = "black", fill=NA, size=0.5),
          axis.line = element_line(size = 0.2, linetype = "solid",
                                   colour = "black"),
          panel.grid = element_line(colour = "gray70", size = 0.2))

```