---
title: "Plasticity - Littorina offspring"
author: "Samuel"
date: "3/25/2020"
output: 
  html_document: 
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

.packages = c("optparse", "dplyr", "tidyr", "rstan", "shinystan", "knitr", "kableExtra", "ggrepel")
# Install CRAN packages (if not already installed)
.inst <- .packages %in% installed.packages()
if(length(.packages[!.inst]) > 0) install.packages(.packages[!.inst])
# Load packages into session
lapply(.packages, require, character.only=TRUE)
```

For now, the data and results belong to the island CZA and phenotype thickness.

```{r, echo=FALSE, warning=FALSE}
island = "CZA"
cz_phen = "mean_thickness"
dat_dir = paste0(island, "_off_SW/", island, "_off_final_data/")
dat_off = read.csv(file = paste0("../../",dat_dir, island, "_off_all_phenos_main_20200110.csv"))
dat_off = separate(data = dat_off, col = "snail_ID", into = c("pop", "ID"), sep = "_")
# table(nchar(as.character(dat_off$ID)))
dat_off[, "generation"] = 1
dat_off[which(nchar(as.character(dat_off$ID)) == 2), "generation"] = 0

dat_gen0 = dat_off[dat_off$generation==0, c("pop", "ID", cz_phen)]
x_meas = aggregate(x = dat_gen0[, cz_phen], by = list(pop = dat_gen0$pop),
                   FUN = function(x) c(mean = mean(x, na.rm = TRUE), sd = sd(x, na.rm = TRUE)))
# mean(dat_gen0[dat_gen0$pop=="B", cz_phen], na.rm = TRUE)
# sd(dat_gen0[dat_gen0$pop=="B", cz_phen], na.rm = TRUE)
x_meas[, paste0("scaled_", cz_phen)] = scale(x_meas$x[,"mean"])[,1]
x_meas[, paste0("scaled_sd_", cz_phen)] = x_meas$x[,"sd"] / sd(x_meas$x[, "mean"])
# xtable::xtable(x = cbind(Pop = x_meas$pop, round(x_meas$x, 2), scaled_mean_thickness = round(x_meas$scaled_mean_thickness, 2),
#                          scaled_sd_thickness = round(x_meas$scaled_sd_mean_thickness, 2)))
# mean(x_meas$x)
# sd(x_meas$x)

dat_gen1 = dat_off[dat_off$generation==1, c("pop", "ID", cz_phen)]
# table(dat_gen1$pop)
# mean(dat_gen1[dat_gen1$pop=="B", cz_phen], na.rm = TRUE)
y_meas = aggregate(x = dat_gen1[, cz_phen], by = list(pop = dat_gen1$pop),
                   FUN = function(y) c(mean = mean(y, na.rm = TRUE), sd = sd(y, na.rm = TRUE)))
y_meas = y_meas[complete.cases(y_meas), ]
y_meas[, paste0("scaled_", cz_phen)] = scale(y_meas$x[,"mean"])[,1]
y_meas[, paste0("scaled_sd_", cz_phen)] = y_meas$x[,"sd"] / sd(y_meas$x[, "mean"])

diff_pop = setdiff(x_meas$pop, y_meas$pop)
x_meas = x_meas[x_meas$pop!=diff_pop,]
# diff_ypop = setdiff(unique(dat_gen1$pop), y_meas$pop)
diff_ypop = which(apply(X = dat_gen1[, c("pop", "ID")], MARGIN = 2,
                        FUN = function(x) !grepl(pattern = "x", x = x)))
dat_gen1 = dat_gen1[diff_ypop, ]
```

For each population it was calculated the mean and standard deviation of the phenotype of interest and these measures were scaled as

\begin{equation}
    \hat{x} = \frac{x - \overline{x}}{\sigma_x}.
\end{equation}

\begin{equation}
    \hat{y} = \frac{y - \overline{y}}{\sigma_y}.
\end{equation}

\begin{equation}
    \hat{\sigma_x} = \frac{\sigma_x}{\sigma_{xt}}.
\end{equation}

\begin{equation}
    \hat{\sigma_y} = \frac{\sigma_y}{\sigma_{yt}}.
\end{equation}

where $x$ and $y$ are the means for one population of parents and offspring, respectively. Scaled variables are marked with $\hat{}$, means are marked with $\overline{}$, and SD of each population and SD for the whole dataset are repsectvely referred as $\sigma$ and $\sigma_t$.

```{r tables, echo=FALSE}
tb1 = cbind(data.frame(table(dat_gen0$pop)[-1]), x_meas$x, x_meas$scaled_mean_thickness, x_meas$scaled_sd_mean_thickness)
colnames(tb1) = c("Pop", "Count", "Mean thick", "SD thick", "Scaled mean", "Scaled SD")
kable(tb1, caption = "Table 1. Observed data for parental populations.",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

tb2 = cbind(data.frame(table(dat_gen1$pop)), y_meas$x, y_meas$scaled_mean_thickness, y_meas$scaled_sd_mean_thickness)
colnames(tb2) = colnames(tb1)
kable(tb2, caption = "Table 2. Observed data for offspring populations.",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r figs, echo=FALSE, fig.cap="Figure 1. Scatter plot of the scaled data."}
ggplot(data = data.frame(x = x_meas$scaled_mean_thickness, y = y_meas$scaled_mean_thickness)) +
    geom_abline(slope = 1, linetype="dashed") +
    geom_point(aes(x = x, y = y), size=2) +
    geom_label_repel(aes(x = x, y = y, label = LETTERS[2:18]),
                     box.padding   = 0.35, 
                     point.padding = 0.5,
                     segment.color = 'grey50') +
    labs(x = "wild sample", y = "lab-reared sample") +
    theme(legend.position = "none",
          # plot.title = element_text(size = 19, hjust = 0.5),
          axis.title = element_text(size = 14),
          axis.text = element_text(size=10),
          axis.ticks = element_line(size = 0.7),
          panel.background = element_blank(),
          panel.border = element_rect(colour = "black", fill=NA, size=0.5),
          axis.line = element_line(size = 0.2, linetype = "solid",
                                   colour = "black"),
          panel.grid = element_line(colour = "gray70", size = 0.2))
```

\

We defined a regression model with measurement error in x an y using Stan.

```{stan output.var="err_in_var", eval=FALSE}
// The input data are vectors 'x', 'sd_x', 'y' and 'sd_y' of length 'N'.
data {
  int<lower=0> N;
  vector[N] x;
  vector<lower=0>[N] sd_x;
  vector[N] y;
  vector<lower=0>[N] sd_y;
}

// The parameters accepted by the model.
parameters {
  vector[N] x_lat;
  vector[N] y_lat;
  real alpha;
  real beta;
  real<lower=0> sigma;
}

transformed parameters {
  vector[N] mu_yhat = alpha + beta * x_lat;
}

// The model to be estimated. We model the output
// 'y_lat' to be normally distributed with mean 'mu_hat = alpha + beta * x_lat'
// and standard deviation 'sigma'.
model {
  x_lat ~ normal(0, 100);
  alpha ~ normal(0, 10);
  beta ~ normal(1, 10);
  sigma ~ normal(0, 100);
  
  x ~ normal(x_lat, sd_x);
  y_lat ~ normal(mu_yhat, sigma);
  y ~ normal(y_lat, sd_y);
}
```

Warning messages:

1: There were 177 divergent transitions after warmup. Increasing adapt_delta above 0.99 may help. See http://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup

2: There were 4 chains where the estimated Bayesian Fraction of Missing Information was low. See
http://mc-stan.org/misc/warnings.html#bfmi-low

\

Since we standardized the data we know $\alpha$ should be centered near 0 and $\beta$ near 1.

```{r stables, echo=FALSE}
res_dir = paste0(island, "_off_SW/", island, "_off_results/")
stbl = read.csv(file = paste0("../../", res_dir, "tables/", island, "_err_in_var_stanfit.csv"))
pars = c("alpha", "beta", "sigma")
pars_idx = which(as.character(stbl$X) %in% pars)
kable(stbl[pars_idx, ], caption = "Table 3. Stan parameter estimates.",
      digits = 2, row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

\

We can the Stan output to diagnose the above sampling problems.

```{r diag, echo=FALSE}
err_in_var = readRDS(paste0("../../", res_dir, "models/", island, "_err_in_var.rds"))
pairs(err_in_var, pars = pars, include = TRUE)
# pairs(err_in_var, pars = "x_lat", include = TRUE)
# pairs(err_in_var, pars = "y_lat", include = TRUE)
# pairs(err_in_var, pars = "mu_yhat", include = TRUE)
```

